---
title: "How big a table can each model handle?"
author: "Marcel Neunhoeffer"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    df-print: kable
vignette: >
  %\VignetteIndexEntry{How big a table can each model handle?}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabfound)

RAM <- c(8, 16, 32, 64) * 1e9
MODELS <- c(
  "TabPFN v2.5" = "tabpfn-v2.5-classifier",
  "TabPFN v2.6" = "tabpfn-v2.6-classifier",
  "TabPFN v3"   = "tabpfn-v3-classifier",
  "TabICL"      = "tabicl-v2-classifier",
  "TabFM"       = "tabfm-1.0.0-classifier",
  "Mitra"       = "mitra-classifier"
)

# One row per model, one column per point on whichever axis is varying.
envelope_table <- function(models = MODELS, n_features = 50, available = RAM,
                           n_query = 200, labels = NULL, ...) {
  labels <- labels %||% paste0(available / 1e9, " GB")
  rows <- lapply(names(models), function(nm) {
    e <- memory_envelope(models[[nm]], n_features = n_features,
                         available = available, n_query = n_query, ...)
    cells <- as.data.frame(as.list(e$max_context), check.names = FALSE)
    names(cells) <- labels
    cbind(data.frame(model = nm), cells)
  })
  do.call(rbind, rows)
}
`%||%` <- function(x, y) if (is.null(x)) y else x
```

Every model here conditions on its training rows at predict time. There
is no fitted parameter vector standing between your table and the
forward pass — the table *is* the input — so its size decides whether
the run happens at all. This vignette is the answer to "will this fit,
and if not, what would?"

Everything below is computed while the vignette is built, from measured
constants, and **needs no weights**: architecture dimensions ship with
the package. You can ask whether a 6.5 GB download is worth making
before you make it.

## Available memory, not installed memory

The tables are indexed by memory **free at the moment you run**, not by
what is on the sticker. The distinction is the whole point. The run that
prompted this work fitted comfortably on a 48 GB machine and was killed
by the operating system the next time, unchanged, because another
process was holding 7 GB.

A rough translation, assuming a machine doing nothing else demanding:

| installed | realistically free |
|---|---|
| 16 GB | 8–11 GB |
| 32 GB | 22–26 GB |
| 64 GB | 50–56 GB |

`estimate_peak_memory()` reports what your machine has free right now, so
you never have to guess:

```{r}
#| label: probe
est <- estimate_peak_memory("tabicl-v2-classifier", n_context = 2000,
                            n_query = 200, n_features = 50)
c(available_gb = round(est$system_avail_bytes / 1e9, 1),
  installed_gb = round(est$system_total_bytes / 1e9, 1))
```

## The headline numbers

Largest training context each model takes **at its own defaults**, with
50 predictor columns and 200 rows to predict, while still leaving the
margin `estimate_peak_memory()` calls comfortable:

```{r}
#| label: headline
envelope_table()
```

Read that as a planning number, not a ceiling. It is the largest table
the estimator would not comment on — roughly half the usable headroom
left unspent, which is what keeps a run alive when something else on the
machine wakes up.

The ceiling itself, the largest table the guard would not refuse:

```{r}
#| label: tight
envelope_table(verdict = "tight")
```

The gap between the two tables is the margin. Running in it is a choice
you can make deliberately — on a machine you control, with nothing else
scheduled, it is often the right one.

Three things stand out, and all three are measured rather than assumed:

* **Nothing is comfortable on 8 GB free.** Before a single activation
  exists there is a fixed cost of 0.6–2.7 GB depending on backend — R,
  libtorch, the allocator's arenas — and 9.5 GB for TabFM. Small tables
  are fine there; fold-sized ones are not, and the ordering between
  models at 8 GB says more about that fixed cost than about the
  architectures.
* **TabPFN v3 goes furthest once there is room to work in** — around
  twenty times TabICL's context and fifty times Mitra's, at 50 columns.
  Not because its state is smaller: v3 carries one vector per *column*
  per row where v2.6 carries one per group of three, and Mitra one per
  feature per row on both axes at once. It reaches further because it is
  the only backend that never holds that state for the whole table —
  `row_chunk_size` runs the first three stages a chunk of rows at a time,
  which turns the term that decides the ceiling from something that grows
  with the table into a constant. See "what the knobs buy".
* **TabFM is in a different regime.** 6.5 GB of weights are resident
  before anything happens, and its 32-member default multiplies what
  comes after.

## Columns cost as much as rows

Capacity is not a row count. Every backend's state is *rows × features*
(or rows × feature-groups), so doubling the columns costs what doubling
the rows costs:

```{r}
#| label: features
cols <- c(10, 25, 50, 100, 200)
envelope_table(models = MODELS[c("TabPFN v2.6", "TabPFN v3", "TabICL",
                                 "Mitra")],
               n_features = cols, available = 32e9,
               labels = paste(cols, "cols"))
```

A 200-column table cuts Mitra's usable context by a factor of 180
relative to 10 columns. If you have wide data and a choice, spend the
effort on feature selection before you spend it on RAM.

Note where v3's advantage runs out. Its chunking bounds the *row* axis,
not the column axis, and its state is a vector per column rather than per
group of three — so at 200 columns TabPFN v2.6 overtakes it. Wide and
short is v2.6's case; tall is v3's.

## Rows you predict are rows too

The forward pass sees the training context *and* the rows being
predicted, so a large test set costs the same per row as a large
training set — up to `predict_chunk_size`, which caps how many are ever
resident at once:

```{r}
#| label: chunk
q <- c(200, 1000, 5000)
data.frame(
  n_query = q,
  max_context = vapply(q, function(n) {
    memory_envelope("tabicl-v2-classifier", n_features = 50,
                    available = 32e9, n_query = n)$max_context
  }, numeric(1))
)
```

Going from 200 to 1,000 rows to predict halves the context you can
afford. Going from 1,000 to 5,000 costs almost nothing more, because
TabICL's default `predict_chunk_size` is 1,024 and only one chunk is ever
resident. This is why `predict()` on a million rows is fine and `fit()`
on a million rows is not.

## What the knobs buy

The single largest lever is the ensemble. Members run one after another,
so the *live* memory is one member's — but the resident high-water mark
climbs anyway, because the allocator takes bigger arenas and does not
give them back. Turning ensembling off is the difference between a run
and no run:

```{r}
#| label: knobs
knobs <- list(
  "defaults (8 members)" = list(),
  "n_estimators = 4"     = list(n_estimators = 4),
  "n_estimators = 1"     = list(n_estimators = 1),
  "defaults + kv_cache"  = list(kv_cache = TRUE)
)
data.frame(
  setting = names(knobs),
  max_context = vapply(knobs, function(o) {
    do.call(memory_envelope,
            c(list("tabicl-v2-classifier", n_features = 50,
                   available = 32e9, n_query = 200), o))$max_context
  }, numeric(1)),
  row.names = NULL
)
```

Note that `kv_cache = TRUE` *costs* capacity. It is a speed setting: it
conditions on the training rows once per `predict()` call instead of once
per chunk per member, and pays for that by holding one cache per member
for the whole call. Reach for it when you are predicting many chunks and
have headroom, not when you are short of memory.

On TabPFN v2.6 and v3 there is a second lever that costs nothing but a
little loop overhead — `save_peak_memory_factor` splits each sublayer's
work, for a bit-identical answer on v2.6 and float32 noise on v3:

```{r}
#| label: spmf
k <- c(1, 2, 4, 8)
data.frame(
  save_peak_memory_factor = k,
  max_context = vapply(k, function(f) {
    memory_envelope("tabpfn-v3-classifier", n_features = 50,
                    available = 32e9, n_query = 200,
                    save_peak_memory_factor = f)$max_context
  }, numeric(1))
)
```

Useful, and free. What it cannot do is change how the peak *grows*: the
tensor it slices is a sublayer's input, not its output, so one copy of
the state survives whatever factor you pick. Measured on v3 at 50
features, the marginal cost of a context row goes from 1.59 MB to
0.549 MB at a factor of 8 — a real 2.9×, and still a straight line.

On v3 alone there is a third lever that bends the line instead of
lowering it. `row_chunk_size` drives the cell embedding, distribution
embedder and column aggregator a chunk of rows at a time, so the
`(rows, columns, embedding)` tensor — 51 KB per row at 100 features — is
never resident whole:

```{r}
#| label: rowchunk
rc <- list("off (one pass)" = NULL, "2048 (the default)" = 2048L,
           "512" = 512L, "128" = 128L)
data.frame(
  row_chunk_size = names(rc),
  max_context = vapply(rc, function(v) {
    memory_envelope("tabpfn-v3-classifier", n_features = 50,
                    available = 32e9, n_query = 200,
                    row_chunk_size = v)$max_context
  }, numeric(1)),
  row.names = NULL
)
```

Two things to read off that. Turning it on is worth about a factor of
ten in rows, which no amount of `save_peak_memory_factor` gets you
because the two bound different tensors. And it *saturates*: below a
chunk of about 512 there is nothing left to win, because what is holding
the ceiling is no longer stages 0-2 but the terms that are linear in the
whole table — the row embeddings, the in-context state, and the KV cache
if you asked for one. Past that point the thing to spend is wall clock,
not chunk size.

It is on by default at the checkpoint's own 2048, which is what the
Python reference does on this architecture and nowhere else. Unlike
`save_peak_memory_factor` it is not bit-identical — it changes the batch
shapes the attention kernel sees, and on the package's 2,664-row parity
fixture it moves the logits by 1.4e-5 of their own scale, which is less
than the reference's own chunked pass moves them.

Because R cannot catch a libtorch allocation failure, there is no
retry-on-OOM to fall back on: the size has to be picked in advance.
`suggest_chunk_sizes()` picks it, searching down from the default and
returning the largest that still fits.

```{r}
#| label: suggest
str(suggest_chunk_sizes("tabpfn-v3-classifier", n_context = 50000,
                        n_features = 50, available = 32e9))
```

## Memory is often not what stops you

Wall clock, measured on the same grid on an Apple-Silicon CPU, at **one
ensemble member** and 90 features. Seconds, by training context size; a
dash means the run died or reached the machine's ceiling before
finishing:

```{r}
#| label: timing
#| echo: false
fs <- list.files(system.file("memory", "measurements", package = "tabfound"),
                 pattern = "json$", full.names = TRUE)
fs <- fs[!startsWith(basename(fs), "ensemble-")]
tm <- do.call(rbind, lapply(fs, function(f) {
  b <- jsonlite::fromJSON(f, simplifyVector = FALSE)
  do.call(rbind, lapply(b$points, function(p) {
    if (!identical(p$status, "ok") || p$n_features != 90) return(NULL)
    data.frame(backend = b$backend, n_context = p$n_context,
               seconds = round(p$elapsed_sec, 1))
  }))
}))
tm <- stats::reshape(tm, idvar = "backend", timevar = "n_context",
                     direction = "wide")
names(tm) <- sub("^seconds\\.", "", names(tm))
tm <- tm[order(tm$backend), ]
rownames(tm) <- NULL
# Blank where the point was not usable: Mitra and TabFM either died at
# 6,400 rows or hit the machine's ceiling on the way.
tm[] <- lapply(tm, function(col) ifelse(is.na(col), "--", as.character(col)))
tm
```

Multiply those by the default member count for the real figure: TabICL's
8 and TabFM's 32. TabFM at 3,200 context rows is 71 seconds for one
member, so around 38 minutes for its default ensemble — long before
memory becomes the binding constraint. Mitra's 145 seconds at the same
size is the steepest curve in the package.

TabPFN and TabICL are seconds; Mitra and TabFM are minutes to hours. If
you are choosing a backend for a resampling loop, that ratio matters more
than the memory table does.

## Choosing

| if your table is | use | why |
|---|---|---|
| under ~2,000 rows, any width | anything | all six are comfortable |
| a few thousand rows, ≤ 100 columns | TabPFN v3 | furthest reach, seconds not minutes |
| tens of thousands of rows | TabPFN v3 | the only backend that bounds the row axis of its embedding tensor; 32,768 × 120 measures 18 GB |
| wide (200+ columns) | TabPFN v2.6, with `save_peak_memory_factor` | v3's chunking bounds rows, not columns, and its state is a vector per column |
| large and you want TabICL's ensembling | TabICL with `n_estimators` lowered | the default 8 costs about three times what one does |
| small (hundreds of rows) | Mitra or TabFM are fine here | this is the size they were published for |

And regardless of size, if the run is going to be tight: subsample the
context. These models do not learn parameters, so halving the context
halves the cost and loses only what those rows contributed as examples —
usually far less than the memory it buys.

## Checking your own case

The tables are generic. Your table is not:

```{r}
#| label: your-case
#| eval: false
# Will this specific run fit, on this machine, right now?
estimate_peak_memory("tabpfn-v3-classifier",
                     n_context = 6426, n_query = 714, n_features = 90)

# How far could I go, with the settings I actually plan to use?
memory_envelope("tabpfn-v3-classifier", n_features = 90,
                available = 24e9, n_estimators = 4,
                save_peak_memory_factor = 4)

# The run is fixed and the memory is what it is: how small do the
# chunks have to be?
suggest_chunk_sizes("tabpfn-v3-classifier", n_context = 2e5,
                    n_features = 50, available = 24e9)
```

You do not have to remember to ask. `fit()` and `predict()` run the same
check themselves and warn before allocating anything, which matters
because running out inside libtorch kills the R process outright — no
error, no traceback, nothing to catch. `options(tabfound.memory_guard =
"error")` turns the warning into a refusal; `"off"` silences it. See
`?estimate_peak_memory`.

## Where the numbers come from

A scaling model whose constants were **measured**: 72 fit/predict runs
across six backends, each in its own process with its peak resident
memory sampled from outside. The method, the raw measurements and the
fitted constants are all in `inst/memory/`.

Two things to know when reading a figure off a table above. The
estimates deliberately err high — on the calibration grid they land
between 1.0× and 1.5× of measured, never below, because for a guard an
over-estimate costs a warning and an under-estimate costs the session.
And they were measured on an Apple-Silicon CPU in float32, which is the
only device calibrated; on another machine, or if a number matters to
you, re-run `inst/memory/calibrate.R`.

The Mitra and TabFM rows are the roughest, since both are large enough
that most of their calibration points ran up against the ceiling of the
machine they were measured on. Read those two as orders of magnitude.
`inst/memory/README.md` has the detail, including where the estimator is
least sure of itself.

If a table here contradicts a run you have actually done, the run is the
better evidence — please open an issue.
