---
title: "Multiple imputation with tabfound"
author: "Marcel Neunhoeffer"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    df-print: kable
vignette: >
  %\VignetteIndexEntry{Multiple imputation with tabfound}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why these models, for this problem

`predict()` throws away everything the model knows except one number.
For most tasks that is the point. For multiple imputation it is exactly
wrong: the spread *between* the imputed data sets is what carries into
the pooled standard error, so a fitted value in every cell would give
you confident nonsense.

Tabular foundation models happen to be unusually well equipped here,
because their heads are distributions to begin with:

| backend | regression head | what a draw is |
|---|---|---|
| `tabpfn` (v2.5, v2.6, v3) | bar distribution over 5000 bins | sample a bin, sample within it |
| `tabicl` | 999-level quantile grid | invert the grid at a uniform |
| `tabfm`, `mitra` | a single number | nothing to draw from |

Every classifier emits a category distribution, which needs no
special-casing at all. `tabfound_impute()` runs chained equations on top
of that.

::: {.callout-note}
The code here is not evaluated when the vignette is built — the weights
are not something to download during `R CMD build`. Every output shown
was produced by running the code exactly as written, against TabPFN
v2.5 on CPU; see [Reproducing this](#reproducing-this).
:::

## The loop

```{r}
#| eval: false
library(tabfound)

mods <- tabfound_models(classifier = "tabpfn-v2.5-classifier",
                        regressor  = "tabpfn-v2.5-regressor")

aq  <- airquality[1:60, ]
imp <- tabfound_impute(aq, m = 5, models = mods, maxit = 3, seed = 20260806)
imp
```

```
tabfound multiple imputation
* 5 imputations, 3 sweeps each
* 2 variables imputed: Ozone and Solar.R
* 29 missing cells of 360
* draw: "auto"
i `as_mids()` for mice, `as_amelia()` for Amelia.
```

What happens per imputation is the textbook chain:

1. every missing cell is filled with a random draw from that column's
   observed values, so the first sweep conditions on something plausible
   rather than on `NA`;
2. each incomplete variable is visited in turn — the model is fitted on
   the rows where that variable is *observed*, with the other columns at
   their current completed values, and new values are drawn for the rows
   where it is missing;
3. that repeats `maxit` times, and the whole thing `m` times.

`tabfound_models()` is a handle, not a model: it loads each checkpoint
at most once, on first use, and reuses it for every variable, sweep and
imputation. A model that is never needed is never loaded, so a data
frame with no incomplete factors does not pay for the classifier.

### Column types pick the head

Exactly as they do for `tabfound()`: numerics go to the regressor,
factors, characters and logicals to the classifier, dates and datetimes
through the regressor and back into their own class.

There is one deliberate exception. **Numeric 0/1 columns with missing
values take the classifier path** and come back as numeric 0/1. A
continuous draw is not a valid value of a dummy, and a downstream
binomial model will refuse it. Amelia treats such columns as continuous;
this is the one place the behaviour was copied on purpose in the other
direction.

### `draw` picks the route

`draw = "auto"` asks the backend for the best it has: `"sample"` for
TabPFN's bar distribution, `"grid"` for TabICL's quantile grid, else
`"quantile"` (an inverse-CDF draw off a fine grid of predicted
quantiles). A backend with only a point estimate is refused rather than
quietly degraded:

```{r}
#| eval: false
tabfound_impute(aq, models = tabfound_models(regressor = "mitra-regressor"))
#> Error: The "mitra" regressor has no predictive distribution to draw from.
#> x It supports only "mean".
#> i Use a backend that samples ("tabpfn") or predicts a quantile grid ("tabicl"),
#>   or pass `draw = "residual"` to draw around the point prediction instead.
```

`draw = "residual"` is available if you insist — a normal draw around
the point prediction sized by the in-sample residuals — but it ignores
everything the predictive distribution knows about skew and
heteroskedasticity, which is why it is opt-in rather than a fallback.

### Reproducibility

The backends sample in torch's RNG, which `set.seed()` does not reach.
Each draw is therefore handed a seed taken from R's stream, so one
`seed =` (or one `set.seed()` beforehand) reproduces the whole
imputation.

## Pooling

`with()` returns a `mira`, mice's container for repeated analyses, so
`pool()` takes it directly — no conversion needed:

```{r}
#| eval: false
library(mice)
summary(pool(with(imp, lm(Ozone ~ Wind + Temp))))
```

```
         term   estimate  std.error  statistic       df      p.value
1 (Intercept) -48.919266 24.0875164 -2.0308970 51.74077 0.0474185087
2        Wind  -0.399109  0.8725424 -0.4574092 18.32698 0.6527528815
3        Temp   1.110571  0.3047787  3.6438597 33.64993 0.0008958215
```

Complete-case analysis on the same 60 rows throws away the 25 with a
missing `Ozone`:

```{r}
#| eval: false
summary(lm(Ozone ~ Wind + Temp, data = aq))$coefficients
```

```
               Estimate Std. Error   t value    Pr(>|t|)
(Intercept) -55.6752128 29.8180173 -1.867167 0.071062489
Wind         -0.6901142  0.8941989 -0.771768 0.445914090
Temp          1.2746429  0.3759286  3.390652 0.001869036
```

Imputation recovers a similar slope on `Temp` from all 60 rows with a
tighter interval (SE 0.305 against 0.376). The imputed rows are
contributing, and the between-imputation spread is already priced into
that standard error — it is not a free lunch, it is 35 rows of
information that deletion was discarding.

### Handing it to mice

For the rest of the mice toolchain — `densityplot()`, `stripplot()`,
`complete()`, anything expecting a `mids`:

```{r}
#| eval: false
mids <- as_mids(imp)
mids$method
#>      Ozone    Solar.R       Wind       Temp      Month        Day
#> "tabfound" "tabfound"         ""         ""         ""         ""

densityplot(mids, ~ Ozone)
mice::complete(imp, "long")   # registered on mice's own generic, no conversion
```

The `mids` is built from a real `mice(maxit = 0)` skeleton rather than
assembled by hand: mice owns the invariants of its own object. The
convergence trace comes with it — `tabfound_impute()` records the mean
and variance of each variable's imputed cells at every sweep, in mice's
own `variable x iteration x imputation` layout — so `plot()` draws the
chains the way it does for a `mice()` fit, and `maxit` is something you
can actually watch converge:

```{r}
#| eval: false
plot(as_mids(imp))
```

### Handing it to Amelia

`mi.meld()` applies the same Rubin's rules and, on this example, returns
the same numbers to the last digit:

```{r}
#| eval: false
library(Amelia)
amp  <- as_amelia(imp)
fits <- lapply(amp$imputations, function(d) lm(Ozone ~ Wind + Temp, data = d))
mi.meld(q  = t(sapply(fits, coef)),
        se = t(sapply(fits, \(f) summary(f)$coefficients[, 2])))
```

```
$q.mi
     (Intercept)      Wind     Temp
[1,]   -48.91927 -0.399109 1.110571

$se.mi
     (Intercept)      Wind      Temp
[1,]    24.08752 0.8725424 0.3047787
```

The `amelia` object's EM-specific slots (`theta`, `mu`, `covMatrices`)
are `NULL`: they describe a multivariate-normal model that was never
fitted here, and filling them with plausible-looking numbers would be
worse than leaving them empty. `mi.meld()`, `missmap()` and the
imputations themselves work; Amelia's EM diagnostics do not.

## Letting mice drive instead

The mirror image: mice runs its own loop and calls in here per variable,
which buys mice's `where`, `blocks`, `post`, `ignore` and convergence
plots.

```{r}
#| eval: false
options(tabfound.models = mods)
imp2 <- mice(aq, method = "tabfound", m = 5, maxit = 3)
```

The trade-off is real and worth knowing: mice hands over an
already-dummy-coded numeric design matrix, so a factor predictor arrives
one column per contrast rather than as the single ordinal column these
networks expect. `tabfound_impute()` keeps factors whole. Prefer it
unless you specifically need something only mice's loop provides.

## Does it actually work?

Unlike everything else in the package this has no reference
implementation to be verified against, so the parity harness has nothing
to say about it. What stands in for parity is a simulation where the
right answer is known by construction — `inst/simulation/run-mi-sim.R`.
Missingness is made to depend on the outcome, which is missing at random
given the data but *not* ignorable for a complete-case analysis, so there
is a real bias for imputation to remove.

```bash
Rscript inst/simulation/run-mi-sim.R --backend=lm --reps=200      # 13s, no weights
Rscript inst/simulation/run-mi-sim.R --backend=tabpfn --reps=200
```

Over 200 replications at `n = 400` with ~42% of a covariate missing:

| arm | bias | coverage |
|---|---|---|
| before deletion | −0.001 | 0.945 |
| complete cases | −0.111 | 0.780 |
| MI, correctly specified imputer | +0.003 | 0.935 |
| MI, **TabPFN** | −0.008 | 0.965 |
| MI, **TabICL** | −0.061 | 0.860 |
| MI, `mice` PMM | +0.013 | 0.945 |

TabPFN removes 93% of the bias and restores coverage, matching both a
correctly specified parametric imputer and `mice`'s PMM without having
been told anything about the data-generating process. Its intervals run
systematically narrower than PMM's, which on one coefficient tips into
undercoverage (0.900 against 0.950) — so PMM remains the safer default
if you want intervals you need not think about.

**TabICL does not do the job.** It removes 45% of the bias here, and on
a binary covariate essentially none. Use the TabPFN backend for
imputation; why TabICL falls short is an open question, and the numbers
are in `inst/simulation/README.md`.

## Properness

A PFN draws each query row independently given a fixed context. Condition
`m` chains on the same observed rows and they differ only in the noise of
the draw: the between-imputation variance carries no *parameter*
uncertainty, which is the defect that makes an imputer improper under
Rubin's rules and makes pooled standard errors too small.

`proper = TRUE` applies the standard correction: resample the observed
rows once per imputation, the nonparametric stand-in for drawing new
parameters, which is what mice's own `norm.boot` / `polyreg.boot` methods
do and what `tabfound_syn()` does by default.

```{r}
#| eval: false
imp_proper <- tabfound_impute(aq, m = 5, models = mods, proper = TRUE)
```

**It is off by default**, which is not what the theory says. The
simulation above was run both ways, on the same 200 data sets:

| | bias `x1` | CI width `x1` | coverage `x2` |
|---|---|---|---|
| complete cases | −0.111 | 0.355 | 0.880 |
| MI, default | −0.008 | 0.314 | 0.900 |
| MI, `proper = TRUE` | **−0.062** | **0.547** | **0.955** |
| MI, `lm` imputer, default | +0.003 | 0.283 | 0.910 |
| MI, `lm` imputer, `proper = TRUE` | +0.003 | 0.325 | 0.935 |

The correction does what the textbook promises for the correctly
specified parametric imputer: coverage moves up, intervals widen 15%,
bias does not move. Applied to TabPFN it fixes coverage the same way and
costs an order of magnitude in bias — most of what the imputation was
worth, since the complete-case bias it exists to remove is −0.111.

The reason is that a bootstrap of the context is not the same model with
different parameters. It is a *worse model*: only 63% of its rows are
distinct, and for a learner whose entire fit is its context, that is a
third of the training data thrown away, with ties the network never saw
during pre-training. The draws come back no wider but less informative —
their correlation with the values that were deleted drops — and that
attenuates every coefficient. Coverage improves because the intervals
grow faster than the error does.

Leave it off to estimate; turn it on if you need nominal coverage more
than you need the point estimate. Whether properness can be had for an
in-context learner without degrading the context is open: every
resampling scheme duplicates rows.

## What this is not

It is a *use* of the fitted models, not a claim about them, and one limit
is worth stating before you rely on it.

**No bounds.** Draws come off the predictive distribution as it is,
tails included. On the example above that is not hypothetical:

```{r}
#| eval: false
oz <- unlist(lapply(imp$imputations, \(d) d$Ozone[is.na(aq$Ozone)]))
c(cells = length(oz), negative = sum(oz < 0), min = min(oz), max = max(oz))
#>    cells negative      min      max
#>      125        3      -16      122
```

Three of 125 imputed ozone concentrations are negative, which is not a
physically possible value. There is no equivalent of Amelia's `bounds`,
and deliberately so — truncating after the fact biases the draws it
keeps. Transform before imputing instead:

```{r}
#| eval: false
aql <- aq; aql$Ozone <- log(aql$Ozone)
impl <- tabfound_impute(aql, m = 5, models = mods, maxit = 3, seed = 20260806)
ozl  <- exp(unlist(lapply(impl$imputations, \(d) d$Ozone[is.na(aq$Ozone)])))
range(ozl)
#> [1]   0.4 168.9
```

Positive by construction, at the cost of imputing on a scale the
analysis model may not want.

## Reproducing this {#reproducing-this}

```{r}
#| eval: false
library(tabfound)
download_model("tabpfn-v2.5-classifier")
download_model("tabpfn-v2.5-regressor")
```

Outputs above are single-threaded CPU on macOS/arm64, R 4.5.2, `torch`
0.16.3, TabPFN v2.5. `mice` and `Amelia` are needed only for the pooling
sections; `tabfound_impute()` itself needs neither.

## See also

* `vignette("tabfound")` — the models, the API, and everything else
* `vignette("synthetic-data", "tabfound")` — the same predictive
  distributions pointed at every cell instead of the missing ones
* `?tabfound_impute`, `?as_mids`, `?as_amelia`,
  `?mice.impute.tabfound`
