---
title: "Synthetic data with tabfound"
author: "Marcel Neunhoeffer"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    df-print: kable
vignette: >
  %\VignetteIndexEntry{Synthetic data with tabfound}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The same models, a different contract

Multiple imputation and synthetic data generation are both fully
conditional specification. They differ all the way down, but the two
rows that carry the weight are *context* and *query*:

| | imputation (`tabfound_impute()`) | synthesis (`tabfound_syn()`) |
|---|---|---|
| cells targeted | the missing ones | every row of every variable |
| conditioning set | all other variables | only variables already visited |
| iteration | `maxit` sweeps toward a stationary distribution | one ordered pass |
| context ("fit") | the current completed data, refit each sweep | the **original** data, fit once per variable |
| query | the current completed covariates | the **synthetic** covariates from earlier in the pass |
| combination rules | Rubin | Reiter / Raab–Nowok–Dibben |

Synthesis factorises the joint as
p(x₁)·p(x₂|x₁)·p(x₃|x₁,x₂)·…, fits each conditional on real data and
draws at synthetic covariates. The first variable is a bootstrap of the
real marginal.

Two consequences follow, and both matter:

* **It is cheaper.** Cost drops from *m* × sweeps × *p* model fits to *p*.
  And because a "fit" here is only conditioning on a context, with
  `proper = FALSE` that context is identical across all *m* syntheses —
  so all *m* share one fit and one forward pass over a stacked query. A
  tree has to be regrown per synthesis; this does not.
* **It is statistically riskier.** Imputation hides two things synthesis
  exposes: the sampler's support behaviour, and the absence of parameter
  uncertainty in row-independent draws. Both get their own section below.

::: {.callout-note}
The code here is not evaluated when the vignette is built. Every output
shown was produced by running the code exactly as written, against
TabPFN v2.5 on CPU; see [Reproducing this](#reproducing-this).
:::

## A first pass

```{r}
#| eval: false
library(tabfound)

mods <- tabfound_models(classifier = "tabpfn-v2.5-classifier",
                        regressor  = "tabpfn-v2.5-regressor")

sds <- tabfound_syn(iris, m = 5, models = mods, seed = 20260807)
sds
```

```
tabfound synthetic data
* 5 synthetic data sets of 150 rows from 150
* 5 variables synthesised: Sepal.Length, Sepal.Width, Petal.Length,
  Petal.Width, and Species
* draw mode: "predictive"
* proper: TRUE
i `as_synds()` for the synthpop toolchain (`compare()`, `utility.gen()`,
  `lm.synds()`).
```

```{r}
#| eval: false
head(sds$syn[[1]], 5)
```

```
  Sepal.Length Sepal.Width Petal.Length Petal.Width    Species
1          5.5    2.615422     4.525384    1.200430 versicolor
2          6.4    2.716248     5.331699    1.573342  virginica
3          6.1    2.789541     3.974613    1.278386 versicolor
4          6.9    3.206068     5.481702    2.237269  virginica
5          7.0    3.202926     4.986701    1.721479 versicolor
```

Two things are visible in that output. `Sepal.Length` holds values with
one decimal — real observed values, because it comes first in the visit
sequence and has nothing to condition on, so it is a bootstrap of its own
marginal. Everything after it carries many decimals: those are genuinely
new numbers off a continuous predictive distribution. `method` records
which is which:

```{r}
#| eval: false
sds$method
#> Sepal.Length  Sepal.Width Petal.Length  Petal.Width       Species
#>     "sample" "regression" "regression" "regression" "classification"
```

Does it keep the structure?

```{r}
#| eval: false
colMeans(iris[1:4])
#> Sepal.Length  Sepal.Width Petal.Length  Petal.Width
#>         5.84         3.06         3.76         1.20
# averaged over the 5 syntheses: 5.92  3.03  3.87  1.23

cor(iris$Petal.Length, iris$Petal.Width)          #> 0.963
# averaged over the 5 syntheses:                  #> 0.955

table(iris$Species)                               #> 50 50 50
# averaged over the 5 syntheses:                  #> 44.2 54.4 51.4
```

And inference, through synthpop's synthetic-data variance estimator:

```{r}
#| eval: false
library(synthpop)
summary(lm.synds(Petal.Width ~ Petal.Length + Species, as_synds(sds)))
```

```
Fit to synthetic data set with 5 syntheses. Inference to coefficients
and standard errors that would be obtained from the original data.

Combined estimates:
                  xpct(Beta) xpct(se.Beta) xpct(z) Pr(>|xpct(z)|)
(Intercept)        -0.041713      0.057936 -0.7200         0.4715
Petal.Length        0.200035      0.035042  5.7085      1.140e-08 ***
Speciesversicolor   0.525878      0.104337  5.0402      4.651e-07 ***
Speciesvirginica    0.919677      0.152033  6.0492      1.456e-09 ***
```

against the real data's `0.230`, `0.435`, `0.838`. Same signs, same
significance pattern, coefficients within about one standard error —
which is roughly what a single-pass synthesiser on 150 rows should be
expected to deliver, and no more.

## The sampler is the main design axis

`synthpop::syn.cart()` draws a leaf, then a donor uniformly from that
leaf's members. TabPFN's regressor is structurally the same two stages —
softmax the bucket logits, pick a bucket, draw uniformly inside it — with
bucket weights in place of leaf membership counts.

The difference propagates everywhere: **CART's support is the observed
values; TabPFN's is the continuum.** `draw_mode` is how far from CART
you want to be, and it is the utility/disclosure trade-off itself, so it
belongs in the write-up rather than in a default nobody reads.

```{r}
#| eval: false
for (dm in c("predictive", "pmm", "rank")) {
  s  <- tabfound_syn(iris, m = 5, models = mods, draw_mode = dm, seed = 20260807)
  sy <- as_synds(s)
  cat(dm,
      "S_pMSE",  mean(utility.gen(sy, iris)$S_pMSE),
      "repU",    mean(colMeans(replicated.uniques(sy, iris)$repU.rm)),
      "novel",   mean(sapply(s$syn, \(d) mean(!(d$Petal.Length %in% iris$Petal.Length)))),
      "\n")
}
```

| mode | `S_pMSE` | replicated uniques | novel `Petal.Length` values |
|---|---:|---:|---:|
| `predictive` | 4.40 | 0.00% | 98.1% |
| `pmm` | 1.76 | 3.33% | 0.0% |
| `rank` | 1.64 | 6.53% | 0.0% |
| `synthpop::syn.cart` | 1.46 | 6.53% | 0.0% |

That table is the whole argument in miniature. `predictive` produces
almost entirely novel values and replicates no unique record — and pays
for it with a propensity score that separates real from synthetic much
more easily (`S_pMSE` of 1 is what an indistinguishable synthesis
scores). `pmm` and `rank` return real donors, which buys back utility and
buys back replication risk in the same move; `rank` reproduces the
marginal exactly, so part of its fidelity is by construction rather than
by merit. `syn.cart` is the incumbent and is right there with them.

::: {.callout-warning}
One dataset, one seed, `n = 150`. This is an illustration of what the
axis *is*, not a benchmark. A frontier worth citing needs several data
sets, several seeds, and a disclosure battery — `replicated.uniques()`
is the weakest of the available measures, not the strongest.
:::

Categorical variables need no such choice: a softmax over the levels plus
one draw per row already *is* the exact analogue of drawing from a leaf's
class proportions.

`clamp` interacts with this and is worth knowing about. It holds
`predictive` draws inside the observed range by flooring and capping
them, which piles the excess mass onto the observed minimum and maximum.
That is a boundary artefact, not a modelled point mass — if a variable
has a real spike at its minimum, say so with `cont_na` (below) rather
than letting `clamp` manufacture one.

## Properness

These models draw each query row independently given a fixed context.
There is no shared drawn θ across rows, so between-synthesis variance
reflects predictive noise and carries **no parameter uncertainty**. That
is the same defect that makes an imputer improper under Rubin's rules —
and for synthesis it is not a robustness check but a validity
requirement, because the Reiter and Raab–Nowok–Dibben estimators that
`lm.synds()` implements assume the mechanism propagates it.

`proper = TRUE` bootstraps the context rows once per synthesis, which is
exactly synthpop's definition of properness. It is the default here, and
it is the one place `tabfound_syn()` deliberately disagrees with
`synthpop::syn()`.

You can see it do its job:

```{r}
#| eval: false
for (p in c(FALSE, TRUE)) {
  s <- tabfound_syn(iris, m = 10, models = mods, proper = p, seed = 11)
  b <- sapply(s$syn, \(d) coef(lm(Petal.Width ~ Petal.Length, d))[["Petal.Length"]])
  cat("proper =", p, " mean", mean(b), " sd", sd(b), "\n")
}
#> proper = FALSE  mean 0.4108  sd 0.0091
#> proper = TRUE   mean 0.4046  sd 0.0141
#> (real data: 0.4158)
```

Both centre on the real slope. The bootstrap widens the spread across
syntheses by about half again — that added spread *is* the parameter
uncertainty, and it is what the variance estimator is expecting to find.
Turning it off costs one fit per variable instead of *m*, so `FALSE` is
the fast setting, not the safe one.

::: {.callout-important}
Whether coverage is actually nominal under the bootstrap is an open
question, not something this vignette answers. The honest experiment is
a calibration study on data-generating processes with known truth,
reporting confidence-interval coverage for a battery of analysis models
under improper and bootstrapped contexts at several *m*. Until that
exists, treat `lm.synds()` intervals from this synthesiser as
provisional.

The imputation side has now run that experiment for *its* use of the same
correction, and the answer there was unwelcome: the bootstrap repaired
coverage and cost an order of magnitude in bias, because a resampled
context keeps only 63% distinct rows and a PFN's fit *is* its context, so
what looks like parameter variation is partly a degraded model. See
`inst/simulation/README.md`; `tabfound_impute()` ships with
`proper = FALSE` as a result.

That does not transfer directly — synthesis needs the properness its
variance estimators assume, in a way estimation does not — but note the
direction in the numbers above: the bootstrapped slope (0.4046) sits
farther from the real one (0.4158) than the improper slope (0.4108) does.
One example at *m* = 10 is not evidence, and it is the direction the
imputation study measured at scale.
:::

## Missing and special values

`NA` is a value to reproduce, not a hole to fill. For a categorical
variable it becomes a level the classifier draws like any other. For a
continuous variable it is modelled in two parts — a categorical model for
*which* special value (or "a real number"), then the regressor for the
real numbers — because a bar distribution cannot put mass on a point.

```{r}
#| eval: false
aq  <- airquality[1:100, ]
s   <- tabfound_syn(aq, m = 5, models = mods, seed = 5)
colMeans(is.na(aq))
#>   Ozone Solar.R    Wind    Temp   Month     Day
#>   0.310   0.070   0.000   0.000   0.000   0.000
# averaged over the 5 syntheses:
#>   0.328   0.076   0.000   0.000   0.000   0.000
```

`cont_na` extends the same machinery to any other point mass — a
zero-inflated income variable, a bottom-coded score:

```{r}
#| eval: false
sds <- tabfound_syn(d, m = 5, models = mods, cont_na = list(income = 0))
```

Without it the spike is treated as a region of a continuous distribution
and gets smeared across its neighbourhood.

## Controlling the pass

**Visit sequence.** With no iteration to wash it out, the order is not a
detail: it decides which conditionals are modelled and which are only
implied. Quantify its effect rather than assuming it away.

```{r}
#| eval: false
tabfound_syn(iris, models = mods,
             visit_sequence = c("Species", "Petal.Length", "Petal.Width",
                                "Sepal.Length", "Sepal.Width"))
```

**Predictors.** A 0/1 matrix in mice's `predictorMatrix` layout.
Entries naming a variable that has not been generated yet are dropped
with a warning rather than silently honoured — a single ordered pass can
only condition on what already exists.

**Methods.** `"regression"`, `"classification"`, `"sample"` (bootstrap
the marginal, no model) or `""` (not synthesised — the real column is
carried through, and is still available to condition on). Useful for a
key you want left alone:

```{r}
#| eval: false
tabfound_syn(d, models = mods, method = c(id = ""))
```

**Size.** `k` sets the rows per synthetic data set independently of the
input, so 150 rows of `iris` can produce 1000:

```{r}
#| eval: false
big <- tabfound_syn(iris, m = 1, k = 1000, models = mods, seed = 6)
table(big$syn[[1]]$Species)
#> 275 364 361
```

Note that `k != nrow(data)` changes which variance estimator applies
downstream.

## Letting synthpop drive instead

`synthpop::syn()` dispatches on method name, so `method = "tabfound"`
hands it the visit sequence and calls in here per variable. That buys
synthpop's `rules`/`rvalues` constraints, its `cont.na` handling,
`k != n`, and — since the object is a native `synds` — everything
downstream with no conversion at all.

```{r}
#| eval: false
options(tabfound.models = mods)
sds <- synthpop::syn(iris, method = "tabfound", m = 2, seed = 20260807)
sds$method
#> Sepal.Length  Sepal.Width Petal.Length  Petal.Width      Species
#>     "sample"   "tabfound"   "tabfound"   "tabfound"   "tabfound"
```

```
  Sepal.Length Sepal.Width Petal.Length Petal.Width    Species
1          6.4    2.787100     5.585445   2.2793370  virginica
2          5.2    2.511311     3.262038   0.8982476 versicolor
3          5.9    2.720448     5.147052   1.7012309  virginica
```

`syn()` refuses any `...` argument that is not named for one of *its own*
methods, so `syn(..., tabfound.draw_mode = "pmm")` cannot work. Tuning
goes through an option instead:

```{r}
#| eval: false
options(tabfound.syn = list(draw_mode = "pmm", donors = 3))
s <- synthpop::syn(iris, method = "tabfound", m = 1, proper = TRUE)
all(s$syn$Sepal.Width %in% iris$Sepal.Width)
#> TRUE
```

Recognised entries are `models`, `draw`, `draw_mode`, `donors`,
`smoothing`, `clamp` and `quantile_grid`. An unrecognised one is an error
rather than a silent no-op.

Two differences from `tabfound_syn()` are worth knowing. synthpop hands
over its own frames, so factor predictors arrive as ordinal codes taken
from the *context's* levels — the query is recoded against them here,
because coding the two independently would make the synthetic covariates
address different categories than the model was fitted on. And `proper`
follows synthpop's default of `FALSE` rather than this package's `TRUE`.

## Evaluating what came out

`tabfound_syn()` results go straight into synthpop's generics, and
`as_synds()` covers the rest:

```{r}
#| eval: false
compare(sds, iris)                              # distributional fidelity
utility.gen(sds, iris)                          # propensity-score pMSE
utility.tab(sds, iris, vars = "Species")        # tabular utility
disclosure(sds, iris,                           # attribute disclosure
           keys = c("Sepal.Length", "Sepal.Width"), target = "Species")

sy <- as_synds(sds)                             # for the non-generic ones
replicated.uniques(sy, iris)
sdc(sy, iris, rm.replicated.uniques = TRUE)
#> no. of replicated uniques removed from each synthetic data set:
#> 0, 0, 0, 0, 0
summary(glm.synds(I(Species == "virginica") ~ Sepal.Width + Sepal.Length,
                  sy, family = binomial))
```

```
Combined estimates:
             xpct(Beta) xpct(se.Beta) xpct(z) Pr(>|xpct(z)|)
(Intercept)   -18.73145       4.03426 -4.6431      3.432e-06 ***
Sepal.Width    -0.47331       0.74865 -0.6322         0.5272
Sepal.Length    3.20019       0.57998  5.5178      3.433e-08 ***
```

A battery worth running before releasing anything: propensity-score
`utility.gen()`, one- and two-way `utility.tab()`, confidence-interval
overlap across a fixed set of analysis models through `lm.synds()` /
`glm.synds()`, marginal and pairwise-correlation recovery, and at least
one three-way interaction — single-pass synthesis is precisely where
higher-order structure degrades. Baselines: `syn.cart` smoothed and
unsmoothed, `syn.norm`, `syn.normrank`.

## Scope

::: {.callout-important}
The context **is** the real data at generation time. This is a fully
conditional synthesiser, not a disclosure-control method, and no
differential-privacy claim is available from it.
:::

Sharp conditionals produce records close to real ones; the `pmm` and
`rank` modes produce records built *from* real values. Measure
replication and attribute disclosure on your own data and report the
frontier — do not infer privacy from the fact that the values are
synthetic.

Two further limits: high-cardinality factors and wide tables run into the
backends' own ceilings rather than anything this code imposes, and there
is no reference implementation to be verified against, so the parity
harness has nothing to say about any of it. Like `tabfound_impute()`,
this is a use of the fitted models, not a claim about them.

## Reproducing this {#reproducing-this}

```{r}
#| eval: false
library(tabfound)
download_model("tabpfn-v2.5-classifier")
download_model("tabpfn-v2.5-regressor")
install.packages("synthpop")
```

Outputs above are single-threaded CPU on macOS/arm64, R 4.5.2, `torch`
0.16.3, `synthpop` 1.9.1, TabPFN v2.5.

## See also

* `vignette("tabfound")` — the models, the API, and everything else
* `vignette("multiple-imputation", "tabfound")` — the same predictive
  distributions pointed at the missing cells instead of every cell
* `?tabfound_syn`, `?as_synds`, `?syn.tabfound`, `?tabfound_synthetic`
