---
title: "Tabular foundation models with tabfound"
author: "Marcel Neunhoeffer"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    df-print: kable
vignette: >
  %\VignetteIndexEntry{Tabular foundation models with tabfound}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this is

Six open-weight tabular foundation models — four distinct architectures,
three generations of one of them — behind one R interface. None needs a
Python runtime at predict time; the forward passes are pure R `torch`.

```{r}
#| eval: false
library(tabfound)
list_backends()
```

```
      name                                              description             parity
1    mitra              Mitra 2-D attention transformer (AutoGluon)     autogluon mitra
2    tabfm TabFM 1.0.0 column/row/ICL transformer (Google Research)        tabfm (PyPI)
3   tabicl        TabICL v2 column/row/ICL transformer (soda-inria)       tabicl (PyPI)
4   tabpfn           TabPFN v2 per-feature transformer (Prior-Labs)       tabpfn (PyPI)
5 tabpfn26                                 TabPFN v2.6 (Prior-Labs) tabpfn 8.2.0 (PyPI)
6  tabpfn3                                   TabPFN v3 (Prior-Labs) tabpfn 8.2.0 (PyPI)
7 tabpfn35                      TabPFN v3.5, multitask (Prior-Labs) tabpfn 9.0.0 (PyPI)
```

The four TabPFN generations are separate backends rather than options on
one, because they are separate architectures: v2.6 moved the
preprocessing inside the network, v3 replaced the decoder as well, and
v3.5 replaced the cell embedder and folded both heads into one
checkpoint.
Each is registered independently and graded independently.

The `parity` column is the point of the package: each backend is checked
end-to-end against the reference Python implementation it was ported
from, stage by stage. See `inst/parity/` for the numbers.

::: {.callout-note}
The code in this vignette is not evaluated when the vignette is built —
the weights range from 40 MB to 6.5 GB and are not something to download
during `R CMD build`. Every output shown was produced by running the
code exactly as written; see [Reproducing this](#reproducing-this).
:::

Two capabilities have vignettes of their own, because they are uses of
these models rather than parts of the prediction API:

* `vignette("multiple-imputation", "tabfound")` — draws from the
  predictive distribution for the missing cells, pooled with `mice` or
  `Amelia`.
* `vignette("synthetic-data", "tabfound")` — the same distributions
  pointed at *every* cell, evaluated with `synthpop`.

## These models don't train

A foundation model for tables does *in-context learning*: `fit()` stores
the training rows, and they are supplied as examples inside the forward
pass at predict time. No gradients, no optimiser, no epochs. That has
three practical consequences worth knowing before you start:

1. Fitting is instant. Predicting is where the compute goes.
2. Every prediction call re-encodes the whole training set, so cost
   scales with `nrow(training data)`, not just with the test set.
3. A "fitted model" is tiny — it is your training data plus a pointer to
   the weights.

## Getting the weights

`download_model()` fetches a model into a local store and, where the
publisher ships a PyTorch pickle, runs the one-time conversion for you:

```{r}
#| eval: false
list_models()          # what is available, and what you already have
download_model("tabpfn-v2.5-classifier")
```

```
i Downloading "tabpfn-v2.5-classifier" from "Prior-Labs/tabpfn_2_5"...
i Converting with `python`...
v "tabpfn-v2.5-classifier" ready in
  '~/Library/Caches/org.R-project.R/R/tabfound/models/tabpfn-v2.5-classifier'
  (42.9 MB).
```

After that the model loads by name — no network, no Python:

```{r}
#| eval: false
fit <- tabfound(Species ~ ., data = iris_train,
                model = "tabpfn-v2.5-classifier")
fit <- tabfound(Species ~ ., data = iris_train, model = "tabpfn")  # family + task
```

A family name plus the task picks the head, so `"tabpfn"` is the
classifier under `tabular_classifier()` and the regressor under
`tabular_regressor()`.

### Versions are named, never inferred

**`"tabpfn"` stays pinned to v2.5.** Adding v2.6 and v3 to the package
does not repoint existing code onto different weights — you ask for a
version by name:

```{r}
#| eval: false
clf <- tabular_classifier("tabpfn-v2.6-classifier")
fit <- tabfound(Species ~ ., data = iris_train, model = "tabpfn-v3")
fit <- tabfound(Species ~ ., data = iris_train, model = "tabpfn-v3.5")
```

### One checkpoint, both tasks

v3.5 is the first generation where the head is chosen per forward pass
rather than baked into the weights. There is no `-classifier` /
`-regressor` pair to choose between: the same id resolves under both
constructors, and `list_models()` reports its task as `both`.

```{r}
#| eval: false
clf <- fit(tabular_classifier("tabpfn-v3.5"), iris[1:4], iris$Species)
reg <- fit(tabular_regressor("tabpfn-v3.5"), iris[2:4], iris$Sepal.Length)
```

Two variants are catalogued by exact id: `"tabpfn-v3.5-fast"` (334 MB
against 876 MB, 8 ICL layers instead of 24 — an alpha checkpoint the
publisher reports as up to 6x faster) and `"tabpfn-v3.5-multiclass"`
(experimental). As with v3's variants they carry no family name, so
`"tabpfn-v3.5"` keeps meaning the default.

v3.5 also brings its estimator's text and date handling. A character
column with more than 30 distinct values becomes 30 LSA features over its
character n-grams, and a `Date` or `POSIXct` becomes calendar features --
on by default for v3.5, opt-in elsewhere with `transform_text = TRUE` and
`transform_dates = TRUE`:

```{r}
#| eval: false
fit <- tabfound(price ~ ., data = listings, model = "tabpfn-v3.5")
fit$expansion      # which columns were expanded, and into how many features
```

v3 publishes six specialised checkpoints alongside its two defaults, and
all eight are catalogued. They are **the same architecture** — identical
config, identical state-dict keys and shapes — so they load through the
same backend with no special handling:

```{r}
#| eval: false
clf <- tabular_classifier("tabpfn-v3-classifier-binary")
reg <- tabular_regressor("tabpfn-v3-regressor-timeseries")
```

| id | what the publisher tuned it for |
|---|---|
| `tabpfn-v3-classifier-binary` | binary targets, <200k rows |
| `tabpfn-v3-classifier-multiclass` | multiclass targets, <200k rows |
| `tabpfn-v3-classifier-ood` | test inputs outside the training distribution |
| `tabpfn-v3-regressor-mediumdata` | <100k rows, alternative preprocessing |
| `tabpfn-v3-regressor-ood` | test inputs outside the training distribution |
| `tabpfn-v3-regressor-timeseries` | synthetic time series |

One of those is worth knowing before you spend the bandwidth. The
**`_ood` checkpoints carry byte-identical weights to the corresponding
defaults** — SHA-256 over the full state dict matches. What makes them
OOD-robust is the preprocessing recipe bundled in the checkpoint, not the
network, and this package reads that recipe from an ensemble dump rather
than from the checkpoint. So a single-pass prediction from
`tabpfn-v3-classifier-ood` is *the same number* as one from
`tabpfn-v3-classifier`. The entries exist so the publisher's names
resolve; they are not distinct models. Family names stay on the defaults:
`"tabpfn-v3"` plus a task never gives you a variant.

::: {.callout-warning}
TabPFN v2.6, v3 and v3.5 weights are released under
`tabpfn-2.6-license-v1.0`, `tabpfn-3-license-v1.0` and
`tabpfn-3-5-license-v1.0`, which permit research, evaluation and
internal benchmarking but **not commercial or production use**. v2.5 is
not under those terms. `list_models()` prints the licence and the
download prompt repeats it — read it before you build on them.
:::

If you skip the download step, the first load asks:

```
! "tabfm-1.0.0-regressor" is not downloaded yet.
* source: "google/tabfm-1.0.0-pytorch"
* size: ~6.6 GB
* licence: NON-COMMERCIAL (weights); Apache-2.0 (source)
* destination: '.../models/tabfm-1.0.0-regressor'
Download it now? (yes/No/cancel)
```

Nothing is fetched without an answer. In a non-interactive session the
load fails with the `download_model()` call to run rather than quietly
pulling gigabytes, and `options(tabfound.download = "always")` opts into
unattended downloads for scripts that want them.

Weights live in `tabfound_home()` —
`tools::R_user_dir("tabfound", "cache")` by default, redirectable with
`options(tabfound.home = )` or `TABFOUND_HOME`. Not inside the installed
package: writing there breaks read-only and shared libraries, and CRAN
forbids it.

### What conversion needs

TabFM and Mitra ship `model.safetensors` + `config.json` directly, so
their download is a plain fetch. **TabPFN and TabICL publish PyTorch
pickles**, which R cannot read — the checkpoints are nested Python dicts
holding a state dict alongside hyperparameters, and R's torch bindings
reject them outright. Converting those needs a Python with `torch` and
`safetensors`, once:

```{r}
#| eval: false
options(tabfound.python = "~/.venvs/tabfound/bin/python")   # or TABFOUND_PYTHON
```

Nothing calls Python again afterwards. The manual route still works if
you prefer to drive it yourself:

```bash
hf download Prior-Labs/tabpfn_2_5 tabpfn-v2.5-classifier-v2.5_default.ckpt --local-dir ckpts
python inst/python/tabpfn_convert_ckpt.py \
  --src ckpts/tabpfn-v2.5-classifier-v2.5_default.ckpt \
  --dst-weights ckpts/converted/tabpfn-v2.5-clf/model.safetensors \
  --dst-config  ckpts/converted/tabpfn-v2.5-clf/config.json \
  --head classifier
```

and then a directory path works anywhere a model name does:

```{r}
#| eval: false
fit <- tabfound(Species ~ ., data = iris_train,
                model = "ckpts/converted/tabpfn-v2.5-clf")
```

You never name the backend. `tabfound()` reads `config.json` and works
out which architecture it is looking at.

## Classification: `iris`

```{r}
#| eval: false
library(tabfound)

set.seed(42)
train <- sort(sample(nrow(iris), 100))
test  <- setdiff(seq_len(nrow(iris)), train)

fit <- tabfound(Species ~ ., data = iris[train, ],
                model = "tabpfn-v2.5-classifier")
fit
```

```
tabfound fit <tabpfn>, mode "classification"
* 4 predictors: Sepal.Length, Sepal.Width, Petal.Length, and Petal.Width
* fitted on 100 rows
* missing values: "pass"
```

`mode` was inferred: `Species` is a factor, so this is classification.
Nothing to configure.

`predict()` returns a tibble, with the column names the tidymodels
conventions use:

```{r}
#| eval: false
head(predict(fit, iris[test, ]), 3)
```

```
# A tibble: 3 x 1
  .pred_class
  <fct>
1 setosa
2 setosa
3 setosa
```

```{r}
#| eval: false
head(predict(fit, iris[test, ], type = "prob"), 3)
```

```
# A tibble: 3 x 3
  .pred_setosa .pred_versicolor .pred_virginica
         <dbl>            <dbl>           <dbl>
1        1.000      0.00000253     0.0000000620
2        1.000      0.000000806    0.0000000306
3        1.000      0.00000273     0.0000000667
```

### Every backend, same call

Only the `model` argument changes:

```{r}
#| eval: false
models <- c(tabpfn  = "tabpfn-v2.5-classifier",
            tabpfn3 = "tabpfn-v3-classifier",
            tabicl  = "tabicl-v2-classifier",
            mitra   = "mitra-classifier",
            tabfm   = "tabfm-1.0.0-classifier")

for (nm in names(models)) {
  fit <- tabfound(Species ~ ., data = iris[train, ], model = models[[nm]])
  acc <- mean(predict(fit, iris[test, ])$.pred_class == iris$Species[test])
  cat(sprintf("%-8s accuracy %.3f\n", nm, acc))
}
```

Measured on this split (100 train / 50 test), CPU, macOS/arm64:

| backend | accuracy | parameters | fit | predict |
|---|---:|---:|---:|---:|
| `tabpfn` (v2.5) | 0.980 | 10.7 M | 0.3 s | 0.11 s |
| `tabpfn3` | 0.980 | 53.2 M | 0.5 s | 0.17 s |
| `tabicl` | 0.960 | 27.6 M | 0.3 s | 0.10 s |
| `mitra`  | 0.960 | 75.7 M | 0.4 s | 0.16 s |
| `tabfm`  | 0.960 | 1639.4 M | 7.1 s | 0.68 s |

"fit" here is almost entirely the cost of loading weights off disk;
TabFM's 7 seconds is 6.5 GB of float32. On a problem this small the
accuracies are one or two test rows apart and the ranking should not be
read as meaningful — `iris` is a 150-row demonstration of the API, not a
benchmark. (`tabpfn26` is not in the table because it was not run for
this vignette; it is graded in `inst/parity/` like the rest.)

## Regression: `cars`

`cars` is 50 observations of stopping distance against speed — one
predictor, and a famously near-linear relationship.

```{r}
#| eval: false
set.seed(42)
ctrain <- sort(sample(nrow(cars), 35))
ctest  <- setdiff(seq_len(nrow(cars)), ctrain)

fit <- tabfound(dist ~ speed, data = cars[ctrain, ],
                model = "tabpfn-v2.5-regressor")
head(predict(fit, cars[ctest, ]), 3)
```

The outcome is numeric, so `mode` is inferred as regression and the
prediction column is `.pred`.

Comparing the four against `lm()`:

| model | RMSE |
|---|---:|
| `lm(dist ~ speed)` | **18.57** |
| `tabpfn` | 19.97 |
| `mitra`  | 20.13 |
| `tabfm`  | 21.35 |
| `tabicl` | 21.41 |

::: {.callout-important}
## The linear model wins, and that is the expected result

`cars` is one numeric predictor with an approximately linear
relationship and 35 training rows. That is precisely the regime a
two-parameter linear model was designed for, and there is nothing for a
foundation model's capacity to buy. These models earn their keep on
wider, messier tables — mixed column types, non-linearities,
interactions, missing values — not on `dist ~ speed`.

`cars` is in this vignette because it is small, familiar, and makes the
API easy to read. Do not read the table as a benchmark.
:::

## Beyond a point prediction

The backends have genuinely different output heads, and the API exposes
that rather than flattening it. This is the section the imputation and
synthesis vignettes build on: everything they do is a draw from one of
these distributions.

**TabPFN** (all three generations) predicts a bar distribution over 5000
bins, so it can give quantiles *and* draw samples:

```{r}
#| eval: false
fit <- tabfound(dist ~ speed, data = cars[ctrain, ],
                model = "tabpfn-v2.5-regressor")
head(predict(fit, cars[ctest, ], type = "quantiles",
             quantiles = c(0.1, 0.5, 0.9)), 3)
```

```
# A tibble: 3 x 3
  .pred_q0.1 .pred_q0.5 .pred_q0.9
       <dbl>      <dbl>      <dbl>
1       6.43       16.4       27.0
2       8.84       18.9       29.8
3      14.0        24.4       36.5
```

```{r}
#| eval: false
draws <- predict(fit, cars[ctest, ], type = "sample",
                 n_samples = 200, seed = 1)
dim(draws)   # 15 x 200 -- one row per test case, one column per draw
```

**TabICL** predicts a 999-level quantile grid directly. `type = "grid"`
hands you the whole thing:

```{r}
#| eval: false
fit <- tabfound(dist ~ speed, data = cars[ctrain, ],
                model = "tabicl-v2-regressor")
dim(predict(fit, cars[ctest, ], type = "grid"))   # 15 x 999
```

**TabFM** and **Mitra** emit a single value per row. Asking them for
quantiles is an error rather than a fabrication:

```{r}
#| eval: false
predict(fit_tabfm, cars[ctest, ], type = "quantiles")
#> Error: The "tabfm" backend has no "quantiles" prediction.
#> i Available: "mean".
```

## Column types and missing values

You do not need to encode anything by hand. Factors, characters,
logicals and dates are all handled, and the encoding is remembered and
re-applied to new data:

```{r}
#| eval: false
set.seed(4)
n <- 160
d <- data.frame(
  num  = rnorm(n),
  grp  = factor(sample(c("a", "b", "c"), n, TRUE)),
  flag = sample(c(TRUE, FALSE), n, TRUE),
  when = as.Date("2021-01-01") + sample(0:400, n, TRUE),
  txt  = sample(c("lo", "hi"), n, TRUE)
)
d$num[sample(n, 12)] <- NA                      # some missing values
d$y <- factor(ifelse(d$num > 0 | d$grp == "a", "yes", "no"))
d$y[is.na(d$y)] <- "no"

fit <- tabfound(y ~ ., data = d[1:120, ], model = "tabpfn-v2.5-classifier")
fit
```

```
tabfound fit <tabpfn>, mode "classification"
* 5 predictors: num, grp, flag, when, and txt
* fitted on 120 rows
* missing values: "pass"
```

Factors become ordinal codes rather than dummy variables — none of these
models one-hot encode features — and dates become their numeric
representation. New data is validated against the training columns, so a
missing column or a novel factor level is reported by name instead of
producing a silently different matrix.

### Missing values differ by backend, and it matters

`na_action` defaults to `"auto"`, which asks the backend what it can
actually handle:

| backend | what the network does with `NA` | `"auto"` resolves to |
|---|---|---|
| `tabpfn`, `tabpfn26`, `tabpfn3`, `tabpfn35` | encodes an explicit is-missing indicator | `pass` |
| `tabfm` | maps it to a sentinel value | `pass` |
| `tabicl` | propagates it — you get `NaN` predictions | `impute` |
| `mitra` | **silently deletes the whole column** | `impute` |

Mitra's behaviour is the one to know about. A single `NA` makes that
column's quantiles all-`NaN`, which sends every value to bucket zero,
which zeroes the column's variance, which the zero-variance guard
flattens. The output is perfectly finite and the feature is gone. This
is the reference implementation's own behaviour, verified against it —
and it is why `tabfound()` imputes for that backend by default rather
than leaving it to you to notice.

If you force the issue, you get told:

```{r}
#| eval: false
tabfound(y ~ ., data = d, model = "tabicl-v2-classifier", na_action = "pass")
#> Warning: The "tabicl" backend has no missing-value handling.
#> x Predictions for rows with "NA" will be "NaN".
#> i Use `na_action = "impute"` or impute before fitting.
```

The built-in imputer is a plain column-mean fill. It is *not* a port of
any reference implementation's imputation and is not covered by the
parity checks — it exists so that backends which cannot see `NA` at all
are still usable. For anything serious, impute deliberately before
fitting.

## Ensembling (TabPFN backends)

A single forward pass is the default, and it is what every number above
was measured with. The reference estimators average several passes over
shuffled and re-preprocessed views of the same data, and `tabpfn`,
`tabpfn26`, `tabpfn3` and `tabpfn35` can all do that here.

A member is described entirely by its config — which primary transform,
whether the transformed block replaces or is appended to the originals,
whether an SVD block and polynomial products are added, the column
shuffle, the class rotation. Build a set in pure R:

```{r}
#| eval: false
X <- as.matrix(iris[train, 1:4]); y <- iris$Species[train]

d <- tempfile(); dir.create(d)
generate_ensemble_configs_native(X, y, n_estimators = 4L, head = "classifier",
                                 variant = "v2.5", output_dir = d)

c1 <- fit(tabular_classifier("tabpfn-v2.5-classifier"), X, y)
c4 <- fit(tabular_classifier("tabpfn-v2.5-classifier",
                             ensemble_configs_dir = d), X, y)

mean(predict(c1, as.matrix(iris[test, 1:4])) == iris$Species[test])   #> 0.98
mean(predict(c4, as.matrix(iris[test, 1:4])) == iris$Species[test])   #> 0.98
```

No change on `iris` — with one test row separating 0.98 from 0.96 there
is nothing for four members to buy. Ensembling earns its keep on harder
problems; this block is here to show the mechanism, not to argue for it.

`variant` has to match your weights: it decides which member menu is
used, and the generations do not share one. The implemented primary
transforms are `none`, `squashing_scaler_default`,
`squashing_scaler_max10`, `quantile_uni`, `quantile_uni_coarse`,
`quantile_uni_fine` and `quantile_uni_extrapolate`, which between them
cover every recipe the eight published v3 checkpoints and both v2
generations ask for. A dump naming anything else raises rather than
substituting a near neighbour.

The native generator is *not* bit-identical to Python at a given seed and
does not try to be — reproducing NumPy's PCG64 in R is not worth the
days. For verified parity, dump the reference's own configs with
`generate_ensemble_configs()` and point at that directory instead; that
is how the ensemble parity is checked.

One catch when combining it with `tabfound()`: the configs record a code
permutation per encoded categorical column, sized from the matrix you
pass in — so that matrix has to be the one the model will actually see.
`tabfound()` molds the frame through hardhat first, which reorders
columns, so generate from the molded matrix rather than the raw frame. A
mismatch is caught rather than silently mis-encoded.

`softmax_temperature` defaults to `0.9`, matching the reference
estimators. Pass `softmax_temperature = 1` for the untempered
distribution.

## Making prediction cheaper

Recall consequence 2 from the top of this vignette: every prediction call
re-encodes the whole training context. That makes cost grow with the
*training* set, and grow again per batch of test rows. Two arguments
change that.

```{r}
#| eval: false
clf <- tabular_classifier("tabpfn-v2.5-classifier",
                          kv_cache = TRUE,             # condition once
                          save_peak_memory_factor = 8) # smaller transients
```

`kv_cache` works on every backend; `save_peak_memory_factor` on
`tabpfn26`, `tabpfn3` and `tabpfn35` only. Asking a backend for a path it does not
have is an error rather than a silent no-op. Both are off by default.

**`kv_cache`** conditions on the training rows once per `predict()` call
and reuses that for every chunk and every ensemble member. On TabPFN the
training rows reach the test rows through exactly two channels — the
key/value projections of the between-items attention, and the
preprocessing statistics fitted on the training rows — and both are
computed once. Measured here on 1500 train / 3000 test rows, 10 columns:

| | no cache | `kv_cache = TRUE` | largest change in `predict_proba` |
|---|---:|---:|---:|
| v2.5 | 2.5 s | **1.3 s** | 0 (bit-identical) |
| v3 | 2.1 s | **1.4 s** | 4.0e-06 |
| TabICL | 4.5 s | **2.1 s** | 2.9e-06 |

The generations and architectures differ in what the cache costs you, and
it is worth knowing which one you are on.

**v2.5 and v3 pay nothing.** Every statistic their networks fit comes
from the training rows alone, so conditioning once cannot answer a
different question. What remains is float32 reduction order — on v2.5
here, not even that.

**TabICL, TabFM and Mitra pay nothing either**, though each holds a
different shape of thing. TabICL slices its context down to the labelled
rows, so it caches one inducing summary per column block plus one
key/value pair per ICL block. TabFM masks instead of slicing and runs the
column stage twice, so it caches two sets of summaries plus its ICL
key/values; masked positions contribute exactly zero, which leaves the
two paths computing the same sum grouped differently — float32 rounding,
around 1e-06 of the logit scale. Mitra's support set is never told the
query rows exist, so its cached predictions are *bit-identical*; the
catch there is memory, because it has no multi-query path and every
attention head has to be kept. `print()` on any of these caches reports
how large it got.

**v2.6 can.** It moved two masks — the constant-column mask and the
within-group informative mask — to a fit over train *and* test. Building
a cache fits them on the training rows alone. Where those agree a cached
prediction is bit-identical; where they don't it is a genuinely different
prediction. The clearest case is a column constant across the training
rows but varying across the test rows: uncached it survives and is used,
cached it was dropped before the cache existed. That is the reference's
own behaviour on the same fixtures, and the parity harness reports it one
by one rather than averaging it away.

**`save_peak_memory_factor`** splits each sublayer's work into that many
chunks, shrinking the attention score matrix each one materialises. It
buys peak memory, not speed — on the run above it cost about 25% in wall
time — and it is bit-identical on v2.6 because it only reorganises work
that was already independent. On v3 it is not quite: chunking changes the
batch shapes the attention kernel sees, and the difference stays at
float32 noise (~4e-06 of the logits' own scale here).

## Multiple imputation and synthetic data

Everything above throws away the model's uncertainty at the last step:
`predict()` returns a number. Two capabilities keep it instead, and each
has its own vignette.

**Multiple imputation** draws each missing cell from the predictive
distribution, so that the spread between the completed data sets carries
into the pooled standard error:

```{r}
#| eval: false
mods <- tabfound_models(classifier = "tabpfn-v2.5-classifier",
                        regressor  = "tabpfn-v2.5-regressor")

imp <- tabfound_impute(airquality[1:60, ], m = 5, models = mods, maxit = 3)
summary(mice::pool(with(imp, lm(Ozone ~ Wind + Temp))))
```

`as_mids()` hands the result to mice, `as_amelia()` to Amelia, and
`mice(method = "tabfound")` runs the other direction. In a simulation
where the truth is known by construction, TabPFN imputation removes 93%
of the bias a complete-case analysis carries and restores nominal
coverage. See `vignette("multiple-imputation", "tabfound")`.

**Synthetic data** points the same distributions at *every* cell:
sequential fully conditional synthesis, each variable modelled on the
real data given the variables before it and drawn at the synthetic values
generated earlier in the pass.

```{r}
#| eval: false
sds <- tabfound_syn(iris, m = 5, models = mods, seed = 20260807)

synthpop::compare(sds, iris)
summary(synthpop::lm.synds(Petal.Width ~ Petal.Length + Species, as_synds(sds)))
```

`as_synds()` opens the whole synthpop toolchain — utility, disclosure and
the synthetic-data variance estimators — and `syn(method = "tabfound")`
lets synthpop drive. The sampler mode and the properness bootstrap are
real decisions with measured consequences; see
`vignette("synthetic-data", "tabfound")`.

Neither has a reference implementation to be verified against, so the
parity harness has nothing to say about either. They are *uses* of the
fitted models, not claims about them.


## Model objects and saving

Fitted models have value semantics, like every other R modelling object.
`fit()` returns a new object and leaves its input alone:

```{r}
#| eval: false
m0 <- tabular_classifier("tabpfn-v2.5-classifier")
m1 <- fit(m0, iris[train, 1:4], iris[train, 5])

is_fitted(m0)   # FALSE -- m0 was not modified
is_fitted(m1)   # TRUE
```

`saveRDS()` does **not** work on these objects. A torch module survives
the round trip structurally, but its tensors come back as dangling
pointers and the failure only appears later, in use. Use the dedicated
functions:

```{r}
#| eval: false
tabfound_save(m1, "model.tabfound")
m2 <- tabfound_load("model.tabfound")
```

The saved file is about **1.3 KB**. Fitting stores context rows rather
than learning parameters, so what gets written is your training data
plus a reference to the weights on disk — not a copy of them. Reloading
re-resolves the weights, so the artifacts (or the HuggingFace cache)
need to still be reachable.

If you do reach for `saveRDS()` anyway, using the result says so:

```{r}
#| eval: false
predict(readRDS("oops.rds"), iris[test, ])
#> Error: This model's weights are no longer available.
#> x Its tensors are dangling pointers, which is what `saveRDS()` leaves
#>   behind: it writes the object's shape but not the weights.
#> i Use `tabfound_save()` and `tabfound_load()` instead.
```

## The engine underneath

`tabfound()` is a layer over a lower-level interface that takes a
numeric matrix directly. Use it when you have already done your own
encoding, or when you want backend-specific options:

```{r}
#| eval: false
clf <- tabular_classifier("tabpfn-v2.5-classifier", device = "cpu")
clf <- fit(clf, as.matrix(iris[train, 1:4]), iris[train, 5])
predict(clf, as.matrix(iris[test, 1:4]), type = "prob")

# Backend-specific arguments pass through:
clf <- tabular_classifier("tabpfn-v2.5-classifier",
                          softmax_temperature = 1)
reg <- tabular_regressor("tabicl-v2-regressor", predict_chunk_size = 256)
```

This layer does no column encoding, no missing-value policy and no
outcome-type inference — it hands your matrix to the network as-is.

## A note on what is and is not ported

Every **network** is complete and verified against its reference, and so
is the **sklearn-style wrapper** around each one: the preprocessing
pipeline, the ensemble of transformed views, and the K/V cache. What is
missing is specific rather than wholesale.

**The three TabPFN backends** have ensembling (feature shuffling, class
rotation, preprocessing variants, SVD and polynomial blocks), the K/V
cache and the chunked forward, all graded — the ensemble against a dumped
reference config directory, the cache and the chunking against both the
reference's own run of the same mechanism and this package's plain
forward.

**TabICL, TabFM and Mitra** have their preprocessing pipelines, their
ensemble generators (Latin-square or random feature permutations,
class shifts, per-member normalisation) and a K/V cache each. What they
do not have: TabFM's heavier `ensemble()` preset (feature crosses, SVD
features, NNLS weights, probability calibration) and its
`TransformToNumerical` datetime handling; TabICL's many-class
hierarchical path; Mitra's fine-tuning, which AutoGluon does by default
and an inference-only package cannot — so Mitra's predictions match a
`MitraClassifier(fine_tune = False)`, not the out-of-the-box one.

## Reproducing this {#reproducing-this}

Every model named in this vignette is fetched the same way. This is the
whole setup:

```{r}
#| eval: false
library(tabfound)

download_model("tabpfn-v2.5-classifier")
download_model("tabpfn-v2.5-regressor")
download_model("tabpfn-v3-classifier")      # non-commercial licence
download_model("tabpfn-v3.5")               # one artifact, both tasks
download_model("tabicl-v2-classifier")
download_model("tabicl-v2-regressor")
download_model("mitra-classifier")
download_model("tabfm-1.0.0-classifier")    # ~6.6 GB

list_models()   # confirm what landed, and where
```

TabPFN and TabICL are converted on the way in, so
`options(tabfound.python = )` has to point at an interpreter with `torch`
and `safetensors` for those two. Skip this block entirely and each model
asks on first use instead.

Timings above are single-threaded CPU on macOS/arm64 with R 4.5.2 and
`torch` 0.16.3, against TabPFN v2.5 and v3. The parity harness that backs
the `parity` column lives in `inst/parity/`; run it with

```bash
Rscript inst/parity/run-parity.R tabpfn
Rscript inst/parity/run-parity.R tabpfn26
Rscript inst/parity/run-parity.R tabpfn3
Rscript inst/parity/run-parity.R tabpfn35
Rscript inst/parity/run-parity.R tabfm
Rscript inst/parity/run-parity.R tabicl
Rscript inst/parity/run-parity.R mitra
```

The two capability vignettes reproduce the same way and add `mice`,
`Amelia` and `synthpop`; neither needs any weights beyond TabPFN v2.5.

## Licences

The package is MIT. The weights are not the package, and carry their own
terms: TabPFN v2.5 and TabICL under their publishers' licences, Mitra
under Apache-2.0, and two sets that are **non-commercial** — TabFM's
weights (separate from its Apache-2.0 source) and **TabPFN v2.6 and v3**,
which permit research, evaluation and internal benchmarking but not
commercial or production use. `list_models()` prints the licence for
each. Check before you deploy.
